Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults

### 2023-03-21 ribose-paper.RMD
updated ribose-paper

Responses to systematic resistance training (RT) vary widely between individuals, with as many as 10-15% showing impaired muscle growth in response to standardized training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. Such inability to benefit from RT may be explained by genetic and/or epigenetic disposition, affecting the internal physiological milieu in terms of ability to mount an anabolic response to RT [@figueiredo_genetic_2021;@thalacker-mercer_cluster_2013].<!-- Check these refs, do they correspond to genet epigen and internal mileu? Kristian: The way I read it, it looks like it corresponds --> Although RT program modifications such as altered exercise volume or intensity may induce more favorable adaptations in some individuals, such benefits are not uniform between individuals [@hammarstrom_benefits_2020;@krieger_single_2009;@schoenfeld_dose-response_2017].<!-- Add ref for more rt program modifications. Kristian: Added krieger et al. 2009 and schoenfeld et al. 2017 --> For instance, although changes in training volume generally induces greater benefits of RT, not all individuals benefit from such modifications [@hammarstrom_benefits_2020]. Therefore, other means than modification of RT variables seems necessary to optimize individual responses.<!-- The above sentence might be a bit contradictory? Im not sure, are you thinking about "other means than modification of RT" being contradictory of "modifications such as altered exercise volume.." sentence? I think there is a point to establishing that yes, increasing exercise volume is a potent way to increase the RT stimulus in general. However, in cases where this alone fails to elicit true responses, other means are necessary. Or am I missinterpreting this? -->

Nutritional supplements such as protein and creatine effectively optimize RT adaptations [@cermak_protein_2012;@lanhers_creatine_2015;@lanhers_creatine_2017;@morton_systematic_2018]. However, it remains unknown if other nutritional adjuvants such as glucose can increase the efficacy of RT. This is surprising, as glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major supplier of energy to cells via ATP synthesis [@mul_exercise_2015;@sebata_control_2018]. Furthermore, energy availability is a decisive factor in *de novo* synthesis of ribosomes [@kusnadi_regulation_2015;@moss_housekeeper_2007;@sebata_control_2018] <!--Add ref that indicate high energy requirement for ribo biog possibly (VONWALDEN19;TANAKA18? MOSS. Kristian: These three should do it--> which in turn determines muscle growth by increasing the muscles translational capacity [@figueiredo_regulation_2019;@hammarstrom_benefits_2020;@sebata_control_2018;@stec_ribosome_2016;@von_walden_ribosome_2019]. Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into a potential additive effect of glucose and RT on ribosome biogenesis.

Transcription of ribosomal RNA (rRNA) by Pol I is considered the limiting step in synthesizing new ribosomes [@moss_promotion_1995].  Multiple signaling pathways converge to regulate rRNA transcription, including the mammalian target of rapamycin (mTORC1) signal-transduction pathway and c-Myc [@kusnadi_regulation_2015;@von_walden_ribosome_2019;@sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. mTOR receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009]. Importantly, mTOR and its downstream target S6K1 serve as a mediator from insulin signaling in skeletal muscle [@hillier_physiological_2000]. The mTORC1 pathway contributes to ribosome biogenesis through the regulation of translation of ribosomal proteins and to forming of the preinitiation complex (PIC) that marks the initiation of rRNA transcription [@figueiredo_regulation_2019;@von_walden_ribosome_2019]. Parallel to mTORC1 [@mori_c-myc_2021; @west_acute_2016], c-Myc increases ribosomal biogenesis directly and through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011]. Additionally, UBF phosphorylation, required for interaction with the rDNA promoter, is increased by high glucose in a mTORC1 dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells[@mariappan_ribosomal_2011]. Independently from UBF and mTORC1, high glucose was shown to lead to chromatin remodeling, which promotes rRNA transcription in cell cultures[@zhai_human_2012]. Together these observations indicate a potential role of glucose in positively affecting muscle rRNA synthesis.

The purpose of this investigation was to test the hypothesis that glucose supplementation before and after<!-- during? Kristian: before and after would be more precise, since they're not strictly ingesting anything during RT - I got a comment on that during the thesis defense --> RT would increase markers of ribosomal abundance following five RT sessions. Secondly we aimed to describe the association between UBF and total RNA abundance in human skeletal muscle. There has been multiple studies the last years suggesting translational capacity to be as important as, if not more, translational efficiency [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022;@figueiredo_revisiting_2019]. Previously, *in vitro* studies has shown UBF and c-Myc to be a important factors in the PIC, c-Myc as a general transcription factor and UBF as a specific transcription factor to rDNA transcription initiation [@mariappan_ribosomal_2011;@von_walden_ribosome_2019]. Coupled with the recent observations by Hammarström et al. [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022], understanding underlying factors regulating and affecting ribosome biogenesis seem key to further our understanding in optimization of RT to individual phenotypes.    
<!-- We should maybe add a line or two about the rational here. Kristian: Tried to add some, what about this? to general? -->

@@

The study was designed as a 12-day double-blinded randomized controlled trial, with a alternating unilateral RT protocol (Fig 1A). Participants were randomly allocated to exercise one leg with glucose and one leg with placebo (1A).
<!--from the "old" methods: either their dominant or non-dominant leg with glucose or placebo before and after RT, and whether they started RT with glucose or placebo. Thus, participants exercised every day, alternating between e.g., one leg + glucose and the other leg + placebo - this might be TMI, as it's presented in the figure as well?--> One person was exclusively responsible for the randomization code and supplement distribution, blinding both investigators and participants in regard to which leg exercised with glucose and placebo. Glucose (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) and placebo (Steviosa, Soma Nordic AS, Oslo, Norway) was masked by mixing with 300ml Fun Light juice/saft (Orkla, Oslo, Norway). A blinded taste test revealed that the participants were not able to disclose the contents of the provided boluses (30g glucosum monohydricum vs. ~0.3g Stevia rebaudiana extract). To ensure equal conditions during training sessions, participants exercised and tested at the same time of day, +/- 1hr (check this) with the same supervisor on pairwise consecutive days. To further standardize this, participants also recorded and repeated their daily macro nutrient intake (protein, fat, carbohydrate) and total calories on pairwise consecutive days.

Six RT sessions were conducted with glucose and six with placebo, allowing a within-subjects analysis of the effects of glucose ingestion before and after RT. Data from the first five RT sessions was used to investigate main outcome measures (total RNA, rRNA and protein) and leg muscle strength, whereas data from the sixth RT session was used to explore secondary outcomes (muscular recovery, plasma glucose levels). Participants were asked to avoid  resistance- or high-intensity training of the legs from Day -7 (Fig 1A) and onwards, until completion of the intervention and post-testing, to ensure the reliability of pre-intervention strength data and minimal interference from other training sources. If participants did conduct exercise training outside of the prescribed protocol, they were asked to ensure equal loading on pairwise consecutive days.

@@

Strength tests were performed prior to (Days -7 and -5, and -1, both legs) and during the intervention (Days 4 and 8 leg 1, 5 and 9 leg two), after session 5 and after finalization of the intervention (Day 11/12 leg 1 and 12/13 leg 2) (Fig 1A). Maximal isometric and isokinetic knee extension torque were measured with a Humac Norm Dynamometer (CSMi, Stoughton, Massechusetts, USA). Individual positions were recorded and standardized from pre-intervention tests (days -7 and -5). Isokinetic peak torque was measured at speeds of 60- and 240 d/s, 2x3 repetitions each, with the first set of each exercise as a sub-maximal warm-up. Isometric peak torque was measured at knee-angle 60, for a maximum of 10 seconds and two repetition per test. The highest peak torque values were used in the final analyses. During days 4, 5, 8 and 9 (Fig 1D, day 4 and 5 = Post 2RT, day 8 and 9 = Post 4RT), humac tests were conducted one hour before RT with the leg performing RT the previous day. During days 11 and 12, humac tests were performed at four time points: I) 45min before RT (1D, Post 5RT), II) 30min after the last RT session (1D, 30 min post 6RT), III) 2hrs after the last RT session (1D, 2h post 6RT), and IV) 23hrs after the last RT session (1D, 23h post 6RT). Test I at day 12/13 included testing of both legs, representing 23hrs post RT session test of one leg and post session 5 test of the other leg.

Assessment of unilateral one repetition maximum (1RM) leg press and knee extension was conducted at pre-intervention testing during familiariztion (days -7 and -5, Fig 1D). The participants performed a general warm-up with 10min cycling on an indoor exercise bicycle. In addition, a protocol consisting of 1x10, 1x6 and 1x3 repetitions, was used as spesific warm-up before each of the tests. All positions were controlled and recorded at the first 1RM test and repeated for the RT sessions. Maximal leg press strength was defined as the maximal load lifted in a controlled fashion, with a knee angle of 45 degrees (start), within a 2.5kg range. Attempts, where participants did not reach 45 degrees during the eccentric phase, were not approved. Maximal knee extension testing followed the same specific warm-up and pause protocol and was defined as maximal load lifted in a controlled fashion, reaching full extension of the knee joint, within a 1.25kg range. Attempts with exaggerated hip movement or beneath full extension were not approved. Two minute breaks were given during specific warm-up, 3min breaks were given between 1RM attempts.

@@

Resistance training consisted of three sets of unilateral leg presses and three sets of knee extensions, with an intensity of 10 repetition maximum (10RM). As a general warm-up, the participants cycled on an indoor exercise bicycle for 5-10min. In addition, before the respective exercises, two 10-repetition warm-up sets were completed at ~50% and ~70% of 10RM. Breaks between working sets was 2min, and progressive loading was used to ensure adequate exercise stimulation throughout the intervention. For safety and standardization purposes, all sessions were monitored by trained personnel. <!-- From the "old" methods: The rate of perceived exertion (RPE, 0–10-point scale) was logged before every session, and the session score was logged 15min after each session - is this necessary info, as we are not reporting these data? --> Lastly, training volume (load and repetitions) was logged for every session.

@@

Two muscle biopsy samples were taken for total RNA extraction per leg per time point, and total RNA was extracted in two duplicates per muscle biopsy. Thus, we had two duplicates per leg per time point, a total of eight RNA samples per participant. Total RNA was extracted using TRIzol according to the manufacturer’s protocol. An exogenous RNA control (λ polyA External Standard Kit, Takara Bio Inc., Shiga, Japan) was added at a fixed amount (0.04 ng ml-1 of Trizol reagent) to enable analysis of target gene expression per unit tissue weight (ELLEFSEN2008,2014). Briefly, muscle tissue was homogenized in TRIzol using 0.5mm RNase-free Zirconium beads (~50 ul) (Next Advanced, Averill Park, NY, USA) and a bullet blender (bb). Chloroform (Sigma-Aldrich, Missouri, USA) was used for phase-separation, and an RNA pellet was precipitated with isopropanol (VWR International, Pennsylvania, USA). For assessment of RNA content and -purity 5µl RNA, from the RNA stock, was eluted in TE-buffer (1:2), for assessment via spectrophotometry. All samples had a 260nm to 280nm ratio > 1.9. The RNA stock was stored at -80°C until further analyses. RNA was reverse transcribed using Super Script IV Reverse Transcriptase (Invitrogen), according to the manufacturer’s instructions. Briefly, 1µl 10 mM dNTP mix, 0.5µl anchored oligo-dT, 0.5µl random hexamer primers (Thermo Scientific), a maximum of 9µl template RNA and 2 µl nuclease-free water were mixed, vortexed and briefly spun down. Samples were then heated at 65°C for 5 minutes, followed by at least 1-minute incubation on ice. For the next step, 4µl 5x SSIV Buffer, 1µl 100 mM DTT, 1 µl RNase OUT, and Super Script IV Reverse Transcriptase (Invitrogen) were mixed, vortexed and briefly spun down and added to the samples. Samples were incubated for 10 minutes at 23°C, 10 minutes at 50-55°C and 10 minutes at 80°C. All samples were reverse transcribed and diluted to 1:50 before quantitative real-time polymerase chain reaction (qPCR). Lambda was used as an external reference gene, added in the RNA extraction (2µl per extraction). qPCR reactions were run on a fast-cycling real-time detection system (Applied Biosystems 7500 fast Real-Time PCR Systems, Life Technologies AS), with a total volume of 10 µl, containing 2 µl of cDNA, gene-specific primers (0.5 µM final concentration) and a commercial master mix (2X SYBR Select Master Mix, Applied Biosystems, Life Technologies AS) (Hammarström et al. 2020). qPCR reactions consisted of 40 cycles (3 s 95°C denaturing and 30 s 60°C annealing) (Hammarström et al. 2020). Raw fluorescence data were exported from the platform-specific software and amplification curves were modelled with a best-fit sigmoidal model using the qPCR package (Ritz & Spiess, 2008) written for R (Hammarström et al. 2020; Team 2018).

@@

A priori power calculations showed that 20 participants would grant a statistical power of 80% (= 0.05), with an expected drop out of 20%. This power calculation was based on an assumption that the effects of glucose ingestion on ribosome biogenesis may equate to the effects of increasing RT volume from low to moderate [@hammarstrom_benefits_2020]. <!-- maybe include the power calculation in extended methods?--> Raw data was imported and analyzed in R via R studios [@Rcore]. Legs were categorized as either glucose or placebo in accordance with the randomization. Before quantitative analyses, total RNA was normalized by wet muscle weight, qPCR values were normalized by wet muscle weight and an external reference gene (Lambda), while quantified protein signals from western blot were normalized by pool signals and total protein signal per gel <!-- add explanations of normalization to extended methods -->. Total RNA and qPCR values were analyzed by comparing post to pre log-fold change score differences. This was achieved by modelling change as response with baseline and supplement (glucose/placebo) as explanatory factors, and mean-centered baseline values as a covariate in a linear mixed effects model using, then calculating the estimated marginal means. Plasma glucose and c-peptide, training volume, and the strength index were analyzed by multiple time-point log-fold change score comparisons, using the same linear mixed effects model with time added as explanatory factor. The linear mixed effects model was designed with the lmer function (Satterthwaite's method) from the lme4 package [@bates_fitting_2014] in R studios, using the lmerTest function to procure p-values. The relationship between <!-- update this --> Log-transformation and mean-centering baseline values was done to control for heteroscedasticity and regression to the mean, respectively. Log-transformed values were reverse-transformed for figure illustration. Values are presented as mean change + 95% CL unless otherwise stated, alpha-levels was set to α = 0.05.

@@

There were no difference in mean change of total training session volume between exercising with placebo or glucose, with an increase of 18 and 16% from session 1 to session 6, respectively (Tab 2; Mean change, 2: *p* = `r pvol.c2`, 3: *p* = `r pvol.c3`, 4: *p* = `r pvol.c4`, 5: *p* = `r pvol.c5`, 6: *p* = `r pvol.c6`). Both exercising with glucose and placebo led to significant increases in mean total training session volume at session four, five and six, compared to session 1 (Tab 2; Mean change, 2: *p* = `r pvol.s2`, 3: *p* = `r pvol.s3`, 4: *p* = `r pvol.s4`, 5: *p* = `r pvol.s5`, 6: *p* = `r pvol.s6`). In regard to the dietary intervention, there were no difference in mean macro nutrient intake (protein, fat, carbohydrates) or total calorie intake between glucose and placebo on pairwise consecutive days (*p* > 0.05).

@@

#### Total RNA and ribosomal RNA

@@

Resistance training with glucose did not induce a higher accumulation of total RNA (fig 2A, *p* = `r trdiffp`) or rRNA  (fig 2B, 47S: *p* = `r rr47diffp`, 18S: *p* = `r rr18diffp`, 28S: *p* = `r rr28diffp`, 5.8S: *p* = `r rr58diffp`, 5S: *p* = `r rr5diffp`) compared to RT with placebo. From baseline to post-intervention, there was a mean increase in levels of total RNA by 26 and 22% after RT with glucose and placebo, respectively (*p* < 0.05, compared to baseline). A robust accumulation was also observed in ribosomal RNA expression in both RT with glucose and placebo, with mean increases between 34-43% (GLU) and 33-41% (PLAC) (*p* < 0.05, compared to baseline) in the four rRNA’s. The expression of the 47S pre-rRNA increased by 37 and 59% with glucose and placebo respectively, where only RT with placebo increased significantly from baseline to post (fig 45, *p* = `r rr47tdiffp`).

#### Protein

>>>>

Responses to systematic resistance training (RT) vary widely between individuals, with as many as 10-15% showing impaired muscle growth in response to standardized training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. Such inability to benefit from RT may be explained by genetic and/or epigenetic disposition, affecting the internal physiological milieu, subsequently affecting the ability to mount an anabolic response to RT [@figueiredo_genetic_2021;@thalacker-mercer_cluster_2013].<!-- Check these refs, do they correspond to genet epigen and internal mileu? Kristian: The way I read it, it looks like it corresponds --> Although RT program modifications such as altered exercise volume or intensity may induce more favorable adaptations in some individuals, such benefits are not uniform between individuals [@hammarstrom_benefits_2020;@krieger_single_2009;@schoenfeld_strength_2017].<!-- Add ref for more rt program modifications. Kristian: Added krieger et al. 2009 and schoenfeld et al. 2017 --> For instance, although changes in training volume generally induces greater benefits of RT, not all individuals benefit from such modifications [@hammarstrom_benefits_2020]. Therefore, other means than modification of RT variables seems necessary to optimize individual responses.<!-- The above sentence might be a bit contradictory? Im not sure, are you thinking about "other means than modification of RT" being contradictory of "modifications such as altered exercise volume.." sentence? I think there is a point to establishing that yes, increasing exercise volume is a potent way to increase the RT stimulus in general. However, in cases where this alone fails to elicit true responses, other means are necessary. Or am I missinterpreting this? -->

Nutritional supplements such as protein and creatine effectively optimize RT adaptations [@cermak_protein_2012;@lanhers_creatine_2015;@lanhers_creatine_2017;@morton_systematic_2018]. However, it remains unknown if other nutritional adjuvants such as glucose can increase the efficacy of RT. This is surprising, as glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major supplier of energy to cells via ATP synthesis [@mul_exercise_2015;@sebata_control_2018]. Furthermore, energy availability is a decisive factor in *de novo* synthesis of ribosomes [@kusnadi_regulation_2015;@moss_housekeeper_2007;@sebata_control_2018] <!--Add ref that indicate high energy requirement for ribo biog possibly (VONWALDEN19;TANAKA18? MOSS. Kristian: These three should do it--> which in turn determines muscle growth by increasing the muscle's translational capacity [@figueiredo_regulation_2019;@hammarstrom_benefits_2020;@sebata_control_2018;@stec_ribosome_2016;@von_walden_ribosome_2019]. Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into a potential additive effect of glucose and RT on ribosome biogenesis.

Transcription of ribosomal RNA (rRNA) by Pol I is considered the limiting step in synthesizing new ribosomes [@moss_promotion_1995].  Multiple signalling pathways converge to regulate rRNA transcription, including the mammalian target of rapamycin (mTORC1) signal-transduction pathway and c-Myc [@kusnadi_regulation_2015;@von_walden_ribosome_2019;@sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. mTOR receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009]. Importantly, mTOR and its downstream target S6K1 serve as a mediator from insulin signalling in skeletal muscle [@hillier_physiological_2000]. The mTORC1 pathway contributes to ribosome biogenesis through the regulation of translation of ribosomal proteins and to forming of the preinitiation complex (PIC) that marks the initiation of rRNA transcription [@figueiredo_regulation_2019;@von_walden_ribosome_2019]. Parallel to mTORC1 [@mori_c-myc_2021; @west_acute_2016], c-Myc increases ribosomal biogenesis directly and through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011]. Additionally, UBF phosphorylation, required for interaction with the rDNA promoter, is increased by high glucose in a mTORC1 dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells[@mariappan_ribosomal_2011]. Independently from UBF and mTORC1, high glucose was shown to lead to chromatin remodelling, which promotes rRNA transcription in cell cultures[@zhai_human_2012]. Together these observations indicate a potential role of glucose in positively affecting muscle rRNA synthesis.

The purpose of this investigation was to test the hypothesis that glucose supplementation before and after<!-- during? Kristian: before and after would be more precise, since they're not strictly ingesting anything during RT - I got a comment on that during the thesis defense --> RT increasea markers of ribosomal abundance following five RT sessions. Secondly we aimed to describe the association between UBF and total RNA abundance in human skeletal muscle. There has been multiple studies the last years suggesting translational capacity to be as important as, if not more, translational efficiency in regard to long-term skeletal muscle adaptations to RT [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022;@figueiredo_revisiting_2019]. Previously, *in vitro* studies has shown UBF and c-Myc to be a important factors in the PIC, c-Myc as a general transcription factor and UBF as a specific transcription factor to rDNA transcription initiation [@mariappan_ribosomal_2011;@von_walden_ribosome_2019]. Coupled with the recent observations by Hammarström et al. [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022], understanding underlying factors regulating and affecting ribosome biogenesis seem key to further our understanding in optimization of RT to individual phenotypes.    
<!-- We should maybe add a line or two about the rational here. Kristian: Tried to add some, what about this? to general? -->

@@

The study was designed as a 12-day double-blinded randomized controlled trial, with an alternating unilateral RT protocol (Fig 1A). Participants were randomly allocated to exercise one leg with glucose and one leg with placebo (1A).
<!--from the "old" methods: either their dominant or non-dominant leg with glucose or placebo before and after RT, and whether they started RT with glucose or placebo. Thus, participants exercised every day, alternating between e.g., one leg + glucose and the other leg + placebo - this might be TMI, as it's presented in the figure as well?--> One person was exclusively responsible for the randomization code and supplement distribution, blinding both investigators and participants in regard to which leg exercised with glucose and placebo. Glucose (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) and placebo (Steviosa, Soma Nordic AS, Oslo, Norway) were masked by mixing with 300ml Fun Light juice/saft (Orkla, Oslo, Norway). A blinded taste test revealed that the participants were not able to disclose the contents of the provided boluses (30g glucosum monohydricum vs. ~0.3g Stevia rebaudiana extract). To ensure equal conditions during training sessions and strength testing, participants exercised and tested at the same time of day, +/- 1hr with the same supervisor on pairwise consecutive days. To further standardize this, participants also recorded and repeated their daily macronutrient intake (protein, fat, carbohydrate) and total calories on pairwise consecutive days.

Six RT sessions were conducted with glucose and six with placebo, allowing a within-subjects analysis of the effects of glucose ingestion before and after RT. Data from the first five RT sessions was used to investigate main outcome measures (total RNA, rRNA and protein) and leg muscle strength, whereas data from the sixth RT session was used to explore secondary outcomes (muscular recovery, plasma glucose and -c-peptide levels). Participants were asked to avoid  resistance- or high-intensity training of the legs from Day -7 (Fig 1A) and onwards, until completion of the intervention and post-testing, to ensure the reliability of pre-intervention strength data and minimal interference from external exercise sources. If participants did conduct exercise training outside of the prescribed protocol, they were asked to ensure equal loading on pairwise consecutive days.

@@

Strength tests were performed prior to (Days -7 and -5, and -1, both legs) and during the intervention (Days 4 and 8 leg 1, 5 and 9 leg two), after session 5 and after finalization of the intervention (Day 11/12 leg 1 and 12/13 leg 2) (Fig 1A). Maximal isometric and isokinetic knee extension torque was measured with a Humac Norm Dynamometer (CSMi, Stoughton, Massechusetts, USA). Individual positions were recorded and standardized from pre-intervention tests (days -7 and -5). Isokinetic peak torque was measured at speeds of 60- and 240 d/s, 2x3 repetitions each, with the first set of each exercise as a sub-maximal warm-up. Isometric peak torque was measured at knee-angle 60, for a maximum of 10 seconds and two repetition per test. The highest peak torque values were used in the final analyses. During days 4, 5, 8 and 9 (Fig 1D, day 4 and 5 = Post 2RT, day 8 and 9 = Post 4RT), humac tests were conducted one hour before RT with the leg performing RT the previous day. During days 11 and 12, humac tests were performed at four time points: I) 45min before RT (1D, Post 5RT), II) 30min after the last RT session (1D, 30 min post 6RT), III) 2hrs after the last RT session (1D, 2h post 6RT), and IV) 23hrs after the last RT session (1D, 23h post 6RT). Test I at day 12/13 included testing of both legs, representing 23hrs post-RT session test of one leg and post-session 5 test of the other leg.

Assessment of unilateral one repetition maximum (1RM) leg press and knee extension was conducted at pre-intervention testing during familiarization (days -7 and -5, Fig 1D). The participants performed a general warm-up with 10min cycling on an indoor exercise bicycle. In addition, a protocol consisting of 1x10, 1x6 and 1x3 repetitions, was used as specific warm-up before each of the tests. All positions were controlled and recorded at the first 1RM test and repeated for the RT sessions. Maximal leg press strength was defined as the maximal load lifted in a controlled fashion, with a knee angle of 45 degrees (start), within a 2.5kg range. Attempts, where participants did not reach 45 degrees during the eccentric phase, were not approved. Maximal knee extension testing followed the same specific warm-up and pause protocol and was defined as maximal load lifted in a controlled fashion, reaching full extension of the knee joint, within a 1.25kg range. Attempts with exaggerated hip movement or beneath full extension were not approved. Two minute breaks were given during specific warm-up, 3min breaks were given between 1RM attempts.

@@

Resistance training consisted of three sets of unilateral leg presses and three sets of knee extensions, with an intensity of 10 repetition maximum (10RM). As a general warm-up, the participants cycled on an indoor exercise bicycle for 5-10min. In addition, before the respective exercises, two 10-repetition warm-up sets were completed at ~50% and ~70% of 10RM. Breaks between working sets were 2min, and progressive loading was used to ensure adequate exercise stimulation throughout the intervention. For safety and standardization purposes, all sessions were monitored by trained personnel. <!-- From the "old" methods: The rate of perceived exertion (RPE, 0–10-point scale) was logged before every session, and the session score was logged 15min after each session - is this necessary info, as we are not reporting these data? --> Lastly, training volume (load and repetitions) was logged for every session.

@@

Two muscle biopsy samples were taken for total RNA extraction per leg per time point, and total RNA was extracted in two duplicates per muscle biopsy. Thus, we had two duplicates per leg per time point, a total of eight RNA samples per participant. Total RNA was extracted using TRIzol according to the manufacturer’s protocol. An exogenous RNA control (λ polyA External Standard Kit, Takara Bio Inc., Shiga, Japan) was added at a fixed amount (0.04 ng ml-1 of Trizol reagent) to enable analysis of target gene expression per unit tissue weight [@ellefsen_improved_2008; @ellefsen_reliable_2014]. Briefly, muscle tissue was homogenized in TRIzol using 0.5mm RNase-free Zirconium beads (~50 ul) (Next Advanced, Averill Park, NY, USA) and a bullet blender (bb). Chloroform (Sigma-Aldrich, Missouri, USA) was used for phase-separation, and an RNA pellet was precipitated with isopropanol (VWR International, Pennsylvania, USA). For assessment of RNA content and -purity 5µl RNA, from the RNA stock, was eluted in TE-buffer (1:2), for assessment via spectrophotometry. All samples had a 260nm to 280nm ratio > 1.9. The RNA stock was stored at -80°C until further analyses. RNA was reverse transcribed using Super Script IV Reverse Transcriptase (Invitrogen), according to the manufacturer’s instructions. Briefly, 1µl 10 mM dNTP mix, 0.5µl anchored oligo-dT, 0.5µl random hexamer primers (Thermo Scientific), a maximum of 9µl template RNA and 2 µl nuclease-free water were mixed, vortexed and briefly spun down. Samples were then heated at 65°C for 5 minutes, followed by at least 1-minute incubation on ice. For the next step, 4µl 5x SSIV Buffer, 1µl 100 mM DTT, 1 µl RNase OUT, and Super Script IV Reverse Transcriptase (Invitrogen) were mixed, vortexed and briefly spun down and added to the samples. Samples were incubated for 10 minutes at 23°C, 10 minutes at 50-55°C and 10 minutes at 80°C. All samples were reverse transcribed and diluted to 1:50 before quantitative real-time polymerase chain reaction (qPCR). Lambda was used as an external reference gene, added in the RNA extraction (2µl per extraction). qPCR reactions were run on a fast-cycling real-time detection system (Applied Biosystems 7500 fast Real-Time PCR Systems, Life Technologies AS), with a total volume of 10 µl, containing 2 µl of cDNA, gene-specific primers (0.5 µM final concentration) and a commercial master mix (2X SYBR Select Master Mix, Applied Biosystems, Life Technologies AS) [@hammarstrom_benefits_2020]. qPCR reactions consisted of 40 cycles (3 s 95°C denaturing and 30 s 60°C annealing) [@hammarstrom_benefits_2020]. Raw fluorescence data were exported from the platform-specific software and amplification curves were modelled with a best-fit sigmoidal model using the qPCR package [@ritz_qpcr_2008] written for R [@hammarstrom_benefits_2020; @Rcore].

@@

A priori power calculations showed that 20 participants would grant a statistical power of 80% (= 0.05), with an expected dropout of 20%. This power calculation was based on an assumption that the effects of glucose ingestion on ribosome biogenesis may equate to the effects of increasing RT volume from low to moderate [@hammarstrom_benefits_2020]. <!-- maybe include the power calculation in extended methods?--> Raw data was imported and analyzed in R via R studios [@Rcore]. Legs were categorized as either glucose or placebo in accordance with the randomization. Before quantitative analyses, total RNA was normalized by wet muscle weight, qPCR values were normalized by wet muscle weight and an external reference gene (Lambda), while quantified protein signals from western blot were normalized by pool signals and total protein signal per gel <!-- add explanations of normalization to extended methods -->. Total RNA and qPCR values were analyzed by comparing post to pre log-fold change score differences. This was achieved by modelling change as response with baseline and supplement (glucose/placebo) as explanatory factors, and mean-centered baseline values as a covariate in a linear mixed effects model using, then calculating the estimated marginal means. Quantified western blot signals and the UBF/total RNA association were analyzed using a linear mixed effects model from the nlme package [@lindstrom_nonlinear_1990], with protein signal per time point as response, with time and supplement as explanatory factors. Plasma glucose and c-peptide, training volume, and the strength index were analyzed by multiple time-point log-fold change score comparisons, using the same linear mixed effects model with time added as explanatory factor. The linear mixed effects model was designed with the lmer function (Satterthwaite's method) from the lme4 package [@bates_fitting_2014] in R studios, using the lmerTest function to procure p-values. Log-transformation and mean-centering baseline values was done to control for heteroscedasticity and regression to the mean, respectively. Log-transformed values were reverse-transformed for figure illustration. Values are presented as mean change + 95% CL unless otherwise stated, alpha-levels was set to α = 0.05.

@@

There was no difference in mean change of total training session volume between exercising with placebo or glucose, with an increase of 18 and 16% from session 1 to session 6, respectively (Tab 2; Mean change, 2: *p* = `r pvol.c2`, 3: *p* = `r pvol.c3`, 4: *p* = `r pvol.c4`, 5: *p* = `r pvol.c5`, 6: *p* = `r pvol.c6`). Both exercising with glucose and placebo led to significant increases in mean total training session volume at session four, five and six, compared to session 1 (Tab 2; Mean change, 2: *p* = `r pvol.s2`, 3: *p* = `r pvol.s3`, 4: *p* = `r pvol.s4`, 5: *p* = `r pvol.s5`, 6: *p* = 0.001). In regard to the dietary intervention, there were no difference in mean macro nutrient intake (protein, fat, carbohydrates) or total calorie intake between glucose and placebo on pairwise consecutive days (*p* > 0.05).

@@

##### Total RNA and ribosomal RNA

@@

Resistance training with glucose did not induce a higher accumulation of total RNA (fig 2A, *p* = `r trdiffp`) or rRNA  (fig 2B, 47S: *p* = `r rr47diffp`, 18S: *p* = `r rr18diffp`, 28S: *p* = `r rr28diffp`, 5.8S: *p* = `r rr58diffp`, 5S: *p* = `r rr5diffp`) compared to RT with placebo. From baseline to post-intervention, there was a mean increase in levels of total RNA by 26 and 22% after RT with glucose and placebo, respectively (*p* < 0.05, compared to baseline). A robust accumulation was also observed in ribosomal RNA expression in both RT with glucose and placebo, with mean increases between 34-43% (GLU) and 33-41% (PLAC) (*p* < 0.05, compared to baseline) in the four rRNAs. The expression of the 47S pre-rRNA increased by 37 and 59% with glucose and placebo respectively, where only RT with placebo increased significantly from baseline to post (fig 45, *p* = `r rr47tdiffp`).

##### Protein
